---
title: "Harbor Seal Contract Report"
author: "Jeff Laake"
date: "9/28/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---



```{r options, results="hide", echo=FALSE}
suppressMessages(knitr::opts_chunk$set(echo = FALSE, results="hide"))
library(pander)
library(xtable)

load(file="gl_bootstraps.rda")
load(file="gl_models.rda")

nreps=length(bs_results)
bspar_N=with(bs_results,sapply(bs_results,function(x)x[[1]]))
llpar_N=apply(bspar_N,1,function(x) sort(x)[nreps*0.025])
ulpar_N=apply(bspar_N,1,function(x) sort(x)[nreps*0.975])

bsMNPL=with(bs_results,sapply(bs_results,function(x)x[[5]]))
llMNPL=apply(bsMNPL,1,function(x) sort(x)[nreps*0.025])
ulMNPL=apply(bsMNPL,1,function(x) sort(x)[nreps*0.975])

bsMNPLratio=with(bs_results,sapply(bs_results,function(x)x[[6]]))
llMNPLratio=apply(bsMNPLratio,1,function(x) sort(x)[nreps*0.025])
ulMNPLratio=apply(bsMNPLratio,1,function(x) sort(x)[nreps*0.975])


```




Final Report for Purchase Order No. 1305M319PNFFS0315 HARBOR SEAL DATA ANALYSIS

# Abstract
Aerial survey count data from 1975-2019 were combined with tidal survey data to provide regional and stock estimates of abundance for 1) Hood Canal stock (HC), 2) Southern Puget Sound stock (SPS), 3) Northern Inland stock (NI), and 4) the Washington portion of the WA/OR coastal stock (C). Generalized logistic population growth curves were fitted to all stocks except for Hood Canal which only had data available from 1991 and there was insufficient growth to fit a curve. The population size at the most recent survey (2019 for SPS and NI and 2014 for C) was above MNPL for these stocks and thus conclude that they are within their OSP range.  No such conclusion can be reached for Hood Canal. Abundance estimates with 95% confidence intervals and NMin values were calculated for inclusion in stock assessment reports. PBR values can be calculated from NMin but were not provided here as they require an agreement on the status of the stocks to set the $F_r$ value. All of the R code and data used for the analysis and result files are available at https://drive.google.com/drive/folders/1HQXJ4HBwHsu040EKQoV9RduHWej1et9T?usp=sharing and will remain there at least through 1 January 2021.


# Data
Aerial survey data from 1975-2019 were gathered from two files provided by NMML and WDFW. The file containing data from 1999-2019 were ready for analysis except for the completion of tidal data at time of the surveys for Hood Canal to construct the abundance estimate based on the London et al correction factor. The historical data from 1975-1998 were in need of filtering and data from 1997 were missing and unavailable.

The R source code files pvCode.r and functions.r create a single data file for harbor seals named pvdf.csv (dataframe pv.df) from the input files OldDataImport.csv (data prior to 1999) and PolygonAerialData.csv (data from 1999-2019). For the historical data there appeared to be duplicate records in the file, so any duplicate records defined by duplicate data for the first 10 fields were removed.  In addition there were errors in the Sitecode values for 6 records, and these are documented in the code with the new replaced values (lines 69-75 in pvCode.r). When the aerial surveys are conducted an estimate is made of the number of seals present.  If for some reason the photograph could not be counted, the estimated count was assigned to the count (lines 91-94 of pvCode.r). Region and Stock fields were not available in the historical data, so they were assigned using the Sitecode, Region and Stock values from the newer data and any missing values were assigned based on parsing the Sitecode (eg 9.x is Eastern Bays) (see lines 97-137 in pvCode.r). That assignment had already been done in the newer data.

For both data sets, counts of seals in the water were removed and any counts in Oregon were removed. The closest tide station was assigned using the file "LocationTideStation.csv" which contained Sitecode and closest tide station.  Any missing values were assigned based on the Sitecode region. The site location, survey time and closest tide station was sent to Josh London with NMML who retrieved tidal data to assign to each count.  The data included nearest_high_height and nearest_low_height (tide height in feet at nearest high and low), tide height at that time and time in minutes from nearest high and from nearest low tide. The data in sequence order (seq) are in Tidesfrom Josh.csv and were appended as columns to the data. Counts at the same location within 50 minutes were combined (see combine.surveytime in functions.r).

The fields in pv.df other than the tidal data were as follows: 1) Sitecode (numeric), 2) Day - date of survey in year month day (read in as character), 3) Survey.time  - survey date and time in POSIXct format, 4) Species - all PV, 5) Count.total - count of all seals, 6) Count.pups - count of pups, 7) Count.nonpups - count of non-pups, 8) Estimated.total - estimated total count at time of survey (not used), 9) Estimated.pups - estimated total pups at time of survey, 10) Region - ten region names within the 4 stocks, 11) Stock - Either Hood Canal, Coastal, Northern Inland or Southern Puget Sound, 12) Tidestation - name of closest tide station, 13) Year - numeric year value, 14) Julian - julian day for survey, 15) year - year as factor variable, 16-20 tidal values. 

For the analysis specific years and range of survey dates were selected. This selection was done in the script data_selection.r which reads in the pvdf.csv file and creates the dataframe pv.df.seldatesyears which is used throughout the analysis code.  The range of julian days used was 198-245 for Northern Inland and Southern Puget Sound stocks, 234-314 for Hood Canal and 121-181 for coastal Washington regions.  Only years with at least one complete survey of each region was used. The selected years by region are contained in the code from lines 21-27 of data_selection.r.   To repeat the analysis it is only necessary to run the script data_selection.r which reads in the pvdf.csv file and then run the analysis scripts described below.


# Methods

## Abundance estimation 
Aerial surveys counted the number of seals ($n$) on land but only some unknown proportion ($p$) were hauled-out on land.  With an estimate of $p$, true abundance $N$ can be estimated as $\hat{N}=n/\hat{p}$ or $\hat{N}=n\hat{f}$ where $\hat{f}=1/\hat{p}$ is often called the correction factor. Both the count ($n$) and $\hat{f}$ are variable. We assumed a binomial distribution for $n$ in the following variance estimator for $\hat{N}$
$$v\hat{a}r(\hat{N})=v\hat{a}r(n)/\hat{p}^2 + n^2v\hat{a}r(\hat{p})/\hat{p}^4=\hat{N}(1-\hat{p})/\hat{p}+\hat{N}^2v\hat{a}r(\hat{p})/\hat{p}^2=\hat{N}(1-\hat{p})/\hat{p}+n^2v\hat{a}r(\hat{f})$$

If the same $\hat{p}$ is used for an estimate at each location and counts are independent then the covariance between estimates $\hat{N}_i$ and $\hat{N}_j$ can be estimated as
$$c\hat{o}v(\hat{N}_i,\hat{N}_j)={n}_i{n}_jv\hat{a}r(\hat{f})$$

The variance-covariance matrix ($\boldsymbol{\Sigma_N}$) in line 21 of SPS&NI Estimate.r and Coastal Estimate.r uses the above 2 formulas and $\hat{f}=1.53$ and $cv(\hat{f})=0.065$ from Huber et al. (2001).

A separate haulout probability model was available from London et al. (2012). The model uses the following variables: 1) hour of day as categorical variable, 2) year as categorical variable for the 3 years study was conducted, 3) number of minutes from closest high tide in 30 categories of 32 minute intervals from -448 to +482, 4) days from August 15 when survey was conducted expressed as a desimal month by dividing by 31 (MFAug15), and 5) the interaction of hour of day with MFAug15. The model is linear on the logit scale and $p$ is computed from the inverse logit $p_i=1/[1+exp(-\beta_i)]$ where $\beta_i$ represents the linear logit value for the $i^{th}$ observation which is a linear combination of the estimated coefficients in the model and the specific covariate values for the observation. We used the models for survey years outside of the study years so we used the average coefficient for the 3 years of the study in computing $\beta_i$. Let $\mathbf{B}$ be the vector of model coefficients and $\mathbf{X}$ be the design matrix for covariates with a row for each observation and a column for each coefficient in the model. The vector $\mathbf{p}$ of haulout probabilities is then expressed as $\mathbf{p}=1/[1+exp(-\mathbf{XB})])$ and the variance-covariance matrix for $p$ is 
$$\boldsymbol{\Sigma_p}=[\mathbf{Xp(1-p)}] \boldsymbol{\Sigma_B} [\mathbf{Xp(1-p)}]' $$
where $\boldsymbol{\Sigma_B}$ is the variance-covariance matrix of the model coefficients. Likewise, the variance-covariance matrix in terms of the vector of correction factors $\mathbf{f}$ is the outer-product ($\odot$) of the $\mathbf{f^2}$ vectors (elementwise squared values) multiplied elementwise with 
$$\boldsymbol{\Sigma_f}=[\mathbf{f^2}\odot \mathbf{f^2}]\circ\boldsymbol{\Sigma_p}$$
which is the cross-product of the $f^2$ vector multiplied element-wise with $\Sigma_p$. The variance-covariance matrix of $\hat{N}$ is 
$$\boldsymbol{\Sigma_N}=[\mathbf{n}\odot \mathbf{n}]\circ\boldsymbol{\Sigma_f} + \mathbf{I}\boldsymbol{\hat{N}}(\mathbf{1}-\mathbf{\hat{p}})/\mathbf{\hat{p}}$$
where $\mathbf{I}$ is a diagonal matrix such that the binomal variance term is only applied to the variance terms.

The fitted model with the coefficients and the variance-covariance matrix was provided by Jay VerHoef in the object all.fit.1.rda. The coefficients are all.fit.1$fixed.effects$estimate and the variance-covariance matrix is all.fit.1$covb.

An estimate of abundance was created for each survey of a site to create a vector $\mathbf{\hat{N}}$ of abundance estimates and a variance-covariance matrix  $\boldsymbol{\Sigma_N}$ for those estimates. Most sites were only surveyed once during a year but some were surveyed multiple times. The mean abundance for the site was used for each site. This was done by creating a matrix $\mathbf{W}$ with a row for each estimate and a column for each survey year. Element $w_{ij}$ of the matrix was 1 if the estimate was for a site that was surveyed only once and it was $1/k$ if the site was surveyed $k$ times in the year. Some sites were not used by seals each year so some weights are 0 if there was no survey of the site in that year. Using the matrix $\mathbf{W}$ simplified replicates (weighting to get mean) and to create a set of annual estimates (sum over the sites in the year) and the variance-covariance matrix of the annual estimates. The annual abundance covariances were not used in this analysis but they are available if needed. The vector of annual abundance estimates $\mathbf{\hat{N}_Y}=\mathbf{\hat{N}}\mathbf{W}$ and the variance-covariance matrix is $\boldsymbol{\Sigma_{N_Y}}=\mathbf{W}'\boldsymbol{\Sigma_N} \mathbf{W}$.

Confidence intervals were constructed based on the log-normal distribution for each annual estimate. The 95% intervals were constructed with $\hat{N}/c,\hat{N}c$ where $c=exp[1.96\sqrt{ln(1+cv(\hat{N})^2)}]$ where $cv(\hat{N})^2=v\hat{a}r(\hat{N})/\hat{N}^2$. Likewise NMin is calculated as $\hat{N}/c$ where $c=exp[0.864\sqrt{ln(1+cv(\hat{N})^2)}]$.

## Population Growth and Status
To evaluate population growth and status within the context of the Marine Mammal Protection Act, we fitted a generalized logistic growth equation to the annual count data by region for each stock following the methods described by Jeffries et al.(2003). For a sequence of counts through time $n_t$ the generalized logistic growth curve equation can be specified as 
$$n_t=n_{t-1}\left[1+R_{max}\left(1-\left({\frac{n_{t-1}}{K}}\right)^z \right)\right]$$
where $R_{max}$ is the maximum growth rate, $K$ is the carrying capacity or asymptotic size, and $z$ is the exponent that adjusts the position of the inflection point of the growth curve. At $z=1$, the inflection point referred to as maximum net productivity level (MNPL) is $K/2$. Optimum sustainable population size is defined as the range of sizes from MNPL to $K$. There are 4 parameters for the equation: 1) $n_0$ which is the initial population size at the starting time, 2) $R_{max}$ the maximum rate of growth achieved at small population sizes, 3) $K$ catrying capacity, and 4) $z$. The maximum likelihood estimation (MLE) method described by Jeffries et al. (2003) assuming proportional errors in the count was used. The MLEs were obtained using the R minimization function nlminb for equation (9) from Jeffries et al.(2003) the summation over the regions used in the analysis. The exponent $z$ is typically very poorly estimated so it was bounded between 1 and 20. Both very small and large values of $z$ create unreasonable growth curve behavior.  The other parameters were bounded as well but the bounds were sufficiently large that they were never imposed on the estimates but can improve optimization behavior.

Growth curve models were fitted to each region because in some years not all regions were counted in a year. This allowed use of all of the available data and the growth curve predictions for the entire stock was computed by summing predictions at each time across the regions within the stock.  A separate initial size $n_0$ and carrying capacity $K$ for each region were included as parameters as there was no reason to expect they would be the same. Whereas, $z$ was assumed to be constant across regions and stocks because it is so poorly estimated with even a large amount of data. However, we did fit 3 models in which $R_{max} was 1) constant (model0), 2) varied by stock (model1), and 3) varied by region (model2). We used the small sample size corrected AICc to choose the best model.

As in Jeffries et al. (2003) we used 1000 parametric bootstraps to compute 95% confidence intervals for the parameters, 95% intervals for the growth curve predictions and for an evaluation of population size at the last survey/MNPL. Values greater than 1 imply the population is within its OSP range. 




# Results
## Growth Curve Model

## Hood Canal Stock

## Southern Puget Sound Stock

## Northern Inland Stock

## Coastal Washington Portion of WA/OR Stock


# Literature Cited

Huber, H.R., S.J. Jeffries, R.F. Brown, R.L. DeLong, and G. VanBlaricom. 2001. Correcting aerial survey of harbor seals (Phoca vitulina richardsi) in Washington and Oregon. Maine Mammal Science 17(2):276-295.

Jeffries, S, H. Huber, J. Calambokidis, and J. Laake.2003. Trends and status of harbor seals in Washington State: 1978-1999. J. Wildl. Manage. 67(1):207-218.

London, J.M., J.M. VerHoef, S. J. Jeffries, M.M. Lance, P.L. Boveng. 2012. Haul-out behavior of harbor seals (Phoca vitulina) in Hood Canal, Washington. PLOS ONE 7(6): e38180. doi:10.1371/journal.pone.0038180


```{r, results='asis'}
cat('\\pagebreak')
```
# Tables

Table 1. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Hood Canal stock from aerial survey data for years between 1991-2019.
```{r,results='asis',echo=FALSE}
HC=read.csv("HCresults.csv")
pander(HC)
```

Table 2. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Southern Puget Sound stock from aerial survey data for years between 1985-2019.
```{r,results='asis',echo=FALSE}
SPS=read.csv("SPSresults.csv")
pander(SPS)
```


```{r, results='asis'}
cat('\\pagebreak')
```
Table 3. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Eastern Bays region of Northern Inland stock from aerial survey data for years between 1983-2019.
```{r,results='asis',echo=FALSE}
EB=read.csv("EBresults.csv")
pander(EB)
```

```{r, results='asis'}
cat('\\pagebreak')
```

Table 4. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for San Juan Island region of Northern Inland stock from aerial survey data for years between 1983-2019.
```{r,results='asis',echo=FALSE}
SJI=read.csv("SJIresults.csv")
pander(SJI)
```
```{r, results='asis'}
cat('\\pagebreak')
```

Table 5. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Strait of Juan de Fuca region of Northern Inland stock from aerial survey data for years between 1983-2019.
```{r,results='asis',echo=FALSE}
SJF=read.csv("SJFresults.csv")
pander(SJF)
```

Table 6. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Northern Inland stock from aerial survey data for years between 1983-2019.
```{r,results='asis',echo=FALSE}
NI=read.csv("NIresults.csv")
pander(NI)
```

```{r, results='asis'}
cat('\\pagebreak')
```
Table 7. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Coastal Estuaries region of coastal stock from aerial survey data for years between 1975-2019.
```{r,results='asis',echo=FALSE}
CE=read.csv("CEresults.csv")
pander(CE)
```

```{r, results='asis'}
cat('\\pagebreak')
```

Table 8. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Outer Coast region of coastal stock from aerial survey data for years between 1980-2014.
```{r,results='asis',echo=FALSE}
OC=read.csv("OCresults.csv")
pander(OC)
```

Table 9. Counts, abundance estimates with standard error, coefficient of variation (CV) and 95% confidence intervals (LCL, UCL) and Nmin value for Washington portion of WA-OR coastal stock from aerial survey data for years between 1980-2014.
```{r,results='asis',echo=FALSE}
C=read.csv("Cresults.csv")
pander(C)
```

```{r, results='asis'}
cat('\\pagebreak')
```


Table 10. Parameter estimates and 95% parametric bootstrap confidence intervals (LCL, UCL). Rmax is $R_m$, n0 is $N_0$ in generalized logistic equation.
```{r, results='asis'}
par_N=glmod1$par
gltab=data.frame(Estimate=par_N,LCL=llpar_N,UCL=ulpar_N)
row.names(gltab)=c("z","Rmax - NI","Rmax - C","Rmax - SPS","n0 -SJF","n0 -SJI","n0 -EB","n0 -CE","n0 -OC","n0 -SPS","K -SJF","K -SJI","K -EB","K -CE","K -OC","K -SPS")
pander(gltab)
```



```{r, results='asis'}
cat('\\pagebreak')
```
# Figures


![](Hood_Canal_Stock_Abundance.pdf)
Figure 1. Hood Canal stock abundance estimates with 95% parametric bootstrap confidence intervals (vertical lines).

```{r, results='asis'}
cat('\\pagebreak')
```

![](Southern_Puget_Sound_Stock_GL.pdf)

Figure 2. Southern Puget Sound stock abundance estimates with 95% parametric bootstrap confidence intervals (vertical lines) with fitted generalized logistic population growth curve (solid curved line) with 95% parametric bootstrap confidence intervals (dashed curved lines) and MNPL estimate (solid horizontal line) and 95% parametric bootstrap confidence intervals (dashed horizontal lines).

![](Northern_Inland_GL.pdf)

Figure 3. Regional counts and fitted generalized logistic population growth curve for EB, SJI and SJF regions in upper panels. Lower panel shows Northern Inland stock abundance estimates with 95% parametric bootstrap confidence intervals (vertical lines) and combined regional fitted generalized logistic population growth curve (solid curved line) with 95% parametric bootstrap confidence intervals (dashed curved lines) and MNPL estimate (solid horizontal line) and 95% parametric bootstrap confidence intervals (dashed horizontal lines).

```{r, results='asis'}
cat('\\pagebreak')
```

![](Coastal_Stock_GL.pdf)

Figure 4. Regional counts and fitted generalized logistic population growth curve for CE and OC regions in upper panels. Lower panel shows Washington coastal stock abundance estimates with 95% parametric bootstrap confidence intervals (vertical lines) and combined regional fitted generalized logistic population growth curve (solid curved line) with 95% parametric bootstrap confidence intervals (dashed curved lines) and MNPL estimate (solid horizontal line) and 95% parametric bootstrap confidence intervals (dashed horizontal lines).





